---
title: "Problem_2"
author: "Ida Hönigmann"
date: '2022-04-06'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Real roots

Let A, B and C be independent random variables, uniformly distributed on (0, 1).

## Calculating the probability

What is the probability that the quadratic equation $Ax^2 + Bx + C = 0$ has real roots?

Using the formula
$$ x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a} $$
we get that $a, b, c$ must satisfy
$$ b^2 - 4ac > 0 \text{ or equivalently } b^2 > 4ac. $$

If we have that $a < \frac{b^2}{4}$ it follows that for all $c \in (0, 1)$ it holds that $ac < \frac{b^2}{4} \iff b^2 > 4ac$.

If $a \geq \frac{b^2}{4}$ we have to have $c < \frac{b^2}{4a}$ in order to satisfy $b^2 > 4ac$.

Using this we can use these limits of integration to get

$$ \mathbb{P}(B^2 > 4AC) =
\int_{0}^{1} (\mathbb{P}(b^2 > 4AC)) db =
\int_{0}^{1} (\int_{0}^{\frac{b^2}{4}}(\mathbb{P}(b^2 > 4aC)) da + \int_{\frac{b^2}{4}}^{1}(\mathbb{P}(b^2 > 4aC)) da) db
$$
$$
= \int_{0}^{1} (\int_{0}^{\frac{b^2}{4}}(1) da + \int_{\frac{b^2}{4}}^{1}(\int_{0}^{\frac{b^2}{4a}}(\mathbb{P}(b^2 > 4ac)) dc) da) db
= \int_{0}^{1} (\frac{b^2}{4} + \int_{\frac{b^2}{4}}^{1}(\int_{0}^{\frac{b^2}{4a}}(1) dc) da) db
$$
$$
= \int_{0}^{1} (\frac{b^2}{4} + \int_{\frac{b^2}{4}}^{1}(\frac{b^2}{4a}) da) db
= \int_{0}^{1} (\frac{b^2}{4} + \frac{b^2}{4} \log(\frac{4}{b^2})) db
= \frac{1}{36}(3\log(4) + 5) \approx 0.2544134
$$

```{r calculation}
1/36 * (3*log(4) + 5)
```

## Estimating the probability

We generate 10000 instances of this problem and check what percentage satisfies the inequality. This gives us an estimate of the probability that $ax^2+bc+c=0$ has real roots.

```{r estimation}
n=10000
a=runif(n)
b=runif(n)
c=runif(n)
sum(b^2>4*a*c)/n
```